\bf Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-04-20 appendix.qmd
Updated Figure 4

For illustration purposes, we only selected cells generated with inDrops ($n=5858$ cells) and UMAP cell embeddings. Also, @xia2023 used first $50$ principal components to generate the UMAP. The objective is to assess the optimized layout by scDEED, and if it does not accurately represent the three clusters with small separations of the PBMC dataset, then select a reasonable \gD{} layout.

@@

error_umap_plot_pbmc <- ggplot(error_pbmc,
                               aes(x = a1,
                                   y = MSE,
                                   colour = method)) +
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8')) +
  scale_y_log10() +
  ylab("log(MSE)") +
  xlab(expression(paste("binwidth (", a[1], ")"))) +
  theme_minimal() +
  theme(panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        legend.position = "none",
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))

@@

```{r}
#| label: published-pbmc-scdeed
#| eval: false

@@

```{r}
#| label: hexbin-pbmc-scdeed
#| eval: false
## Compute hexbin parameters
num_bins_x_pbmc <- 12
lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "pc"
)

@@

```{r}
#| label: pbmc-umap-model-proj-scdeed
#| eval: false

@@

```{r}
#| echo: false
#| eval: false
#| fig-cap: "Model in \\gD{}, on the layout a, and two views of the fit in projections from $50\\text{-}D$, for the PBMC data ($n =  5858$ and $p = 50$). Layout shows close clusters. In $50\\text{-}D$, the data shows "
#| label: fig-model-pbmc-author-proj-scdeed
#| fig-pos: H
#| fig-width: 15
#| fig-height: 5
#| out-width: 30%

>>>>

For illustration purposes, we only selected cells generated with inDrops ($n=5858$ cells) and UMAP cell embeddings. Also, @xia2023 used first $9$ principal components to generate the UMAP. The objective is to assess the optimized layout by scDEED, and if it does not accurately represent the three clusters with small separations of the PBMC dataset, then select a reasonable \gD{} layout.

@@

error_umap_plot_pbmc <-  plot_mse(error_pbmc) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19, 20)]) +
  scale_color_manual(values=c('#e41a1c','#377eb8'))
```

@@

```{r}
#| label: published-pbmc-scdeed

@@

```{r}
#| label: hexbin-pbmc-scdeed

## Compute hexbin parameters
num_bins_x_pbmc <- 12
lim1 <- pbmc_scaled_obj_umap$lim1
lim2 <- pbmc_scaled_obj_umap$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
    highd_data = training_data_pbmc,
    nldr_data = umap_pbmc_scaled,
    bin1 = num_bins_x_pbmc,
    r2 = r2_pbmc,
    q = 0.1,
    is_bin_centroid = TRUE
  )

@@

```{r}
#| label: pbmc-umap-model-proj-scdeed

@@

```{r}
#| fig-cap: "Model in \\gD{}, on the layout a, and two views of the fit in projections from $50\\text{-}D$, for the PBMC data ($n =  5858$ and $p = 50$). Layout shows close clusters. In $50\\text{-}D$, the data shows "
#| label: fig-model-pbmc-author-proj-scdeed
#| fig-pos: H
#| fig-width: 15
#| fig-height: 5
#| out-width: 30%

### 2025-04-20 appendix.tex
Updated Figure 4

For illustration purposes, we only selected cells generated with inDrops
(\(n=5858\) cells) and UMAP cell embeddings. Also, \citet{xia2023} used
first \(50\) principal components to generate the UMAP. The objective is
to assess the optimized layout by scDEED, and if it does not accurately
represent the three clusters with small separations of the PBMC dataset,
then select a reasonable \gD{} layout.

@@

>>>>

For illustration purposes, we only selected cells generated with inDrops
(\(n=5858\) cells) and UMAP cell embeddings. Also, \citet{xia2023} used
first \(9\) principal components to generate the UMAP. The objective is
to assess the optimized layout by scDEED, and if it does not accurately
represent the three clusters with small separations of the PBMC dataset,
then select a reasonable \gD{} layout.

@@

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{appendix_files/figure-pdf/fig-model-pbmc-author-proj-scdeed-1.pdf}

\caption{\label{fig-model-pbmc-author-proj-scdeed}Model in \gD{}, on the
layout a, and two views of the fit in projections from \(50\text{-}D\),
for the PBMC data (\(n =  5858\) and \(p = 50\)). Layout shows close
clusters. In \(50\text{-}D\), the data shows}
